\LARGE \bf On the Commissioning of the 12 GeV HMS Drift Chambers, Electronics/Computer Dead Time Monitoring and Overview of the D(e,e'p)n Experimental Run Plan

### 2017-10-22 NRF_Final_Report/report.tex
started 2nd page of NRF report

\documentclass[letterpaper, 10 pt, conference]{ieeeconf}  % Comment this line out
                                                          % if you need a4paper
%\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4
                                                          % paper

\IEEEoverridecommandlockouts                              % This command is only
                                                          % needed if you want to
                                                          % use the \thanks command
\overrideIEEEmargins
% See the \addtolength command later in the file to balance the column lengths
% on the last page of the document

% The following packages can be found on http:\\www.ctan.org
%\usepackage{graphics} % for pdf, bitmapped graphics files
%\usepackage{epsfig} % for postscript graphics files
%\usepackage{mathptmx} % assumes new font selection scheme installed
%\usepackage{times} % assumes new font selection scheme installed
%\usepackage{amsmath} % assumes amsmath package installed
%\usepackage{amssymb}  % assumes amsmath package installed

\title{\LARGE \bf
On the Commissioning of the 12 GeV HMS Drift Chambers, Electronics/Computer Live Time Monitoring and D(e,e'p)n Experimental Run Plan
}

%\author{ \parbox{3 in}{\centering Huibert Kwakernaak*
%         \thanks{*Use the $\backslash$thanks command to put information here}\\
%         Faculty of Electrical Engineering, Mathematics and Computer Science\\
%         University of Twente\\
%         7500 AE Enschede, The Netherlands\\
%         {\tt\small h.kwakernaak@autsubmit.com}}
%         \hspace*{ 0.5 in}
%         \parbox{3 in}{ \centering Pradeep Misra**
%         \thanks{**The footnote marks may be inserted manually}\\
%        Department of Electrical Engineering \\
%         Wright State University\\
%         Dayton, OH 45435, USA\\
%         {\tt\small pmisra@cs.wright.edu}}
%}

@@

\twocolumn[
  \begin{@twocolumnfalse}
    \maketitle
    \begin{abstract}
      Three separate topics, all of equal importance, are briefly discussed. The new (12 GeV Era) HMS Drift Chambers are ready to be put in the HMS
      detector stack, in place of the old HMS Chambers. Several efficcinecy tests were performed on one of the chambers during the second week of October
      2017. The efficcinecies were determined to be better than 99$\%$. The second chamber has not been tested yet, but it is
      expected to behave the same since both chambers were tested under similar conditions in the past. Live time studies are currently in progress to
      determine how many physics events (triggers) are actually lost due to computer and electronic deadtime inherent in our experimental equipment. 
      There had been some technical issues found related to these studies that are being addressed by the Jefferson Lab DAQ group. The experimental run
      plan of my thesis experiment, the electro-disintegration of deuteron (D(e,e'p)n), is briefly discussed as there has been some small changes in
      the kinematic settings. \\

    \end{abstract}
  \end{@twocolumnfalse}
  ]

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{INTRODUCTION}

This template, modified in MS Word 2003 and saved as ÒWord 97-2003 \& 6.0/95 Ð RTFÓ for the PC, provides authors with most of the formatting specifications needed for preparing electronic versions of their papers. All standard paper components have been specified for three reasons: (1) ease of use when formatting individual papers, (2) automatic compliance to electronic requirements that facilitate the concurrent or later production of electronic products, and (3) conformity of style throughout a conference proceedings. Margins, column widths, line spacing, and type styles are built-in; examples of the type styles are provided throughout this document and are identified in italic type, within parentheses, following the example. Some components, such as multi-leveled equations, graphics, and tables are not prescribed, although the various table text styles are provided. The formatter will need to create these components, incorporating the applicable criteria that follow.

\section{PROCEDURE FOR PAPER SUBMISSION}

\subsection{Selecting a Template (Heading 2)}

First, confirm that you have the correct template for your paper size. This template has been tailored for output on the US-letter paper size. Please do not use it for A4 paper since the margin requirements for A4 papers may be different from Letter paper size.

>>>>

\documentclass[letterpaper, 10 pt, conference]{ieeeconf}  % Comment this line out
                                                          % if you need a4paper
%\documentclass[a4paper, 10pt, conference]{ieeeconf}      % Use this line for a4
\usepackage{url}
\usepackage[font=footnotesize,labelfont=bf]{caption}
\usepackage{graphicx}
\usepackage[scientific-notation=true]{siunitx}
\usepackage{subcaption}
\usepackage{setspace}
\usepackage[top=0.6 in, bottom=0.7 in, left=0.7in, right=0.7in]{geometry}
\usepackage{blindtext}
\usepackage[utf8]{inputenc}
\usepackage{setspace}
\usepackage{vector}
\usepackage{amsmath}
\usepackage{empheq}
\usepackage{amssymb}
\usepackage{tgtermes}
\usepackage{gensymb}
\DeclareGraphicsExtensions{.pdf,.png,.jpg}
\usepackage{epstopdf}
\usepackage[singlelinecheck=false]{caption}
\usepackage{datetime}
\usepackage[hyperfootnotes=false, pdfnewwindow, colorlinks=true]{hyperref}

\AtBeginDocument{% optionally set colors to your liking
  \hypersetup{
    urlcolor=red,
    citecolor=green,
    linkcolor=blue,
  }%
}                                                         % paper

\IEEEoverridecommandlockouts                              % This command is only
                                                          % needed if you want to
                                                          % use the \thanks command
\overrideIEEEmargins

\title{\LARGE \bf
  On the Commissioning of the 12 GeV HMS Drift Chambers, Electronics/Computer Dead Time Monitoring
  and Overview of the D(e,e'p)n Experimental Run Plan}

@@

\twocolumn[
  \begin{@twocolumnfalse}
    \maketitle
    \begin{abstract}
      Three separate topics, all of equal importance, are briefly discussed. The new (12 GeV Era) HMS Drift Chambers are ready to be put in the HMS
      detector stack, in place of the old HMS Chambers. Several efficcinecy tests were performed on one of the chambers during the second week of October
      2017. The efficciecies were determined to be better than 99$\%$. The second chamber has not been tested yet, but it is
      expected to behave the same since both chambers were tested under similar conditions in the past. Dead time studies are currently in progress to
      determine how many physics events (triggers) are actually lost due to computer and electronic deadtime inherent in our experimental equipment. 
      There had been some technical issues found related to the computer livetime that are being addressed by the Jefferson Lab DAQ group. The experimental run
      plan of my thesis experiment, the electro-disintegration of Deuteron, is briefly discussed as the kinematics have slightly changed and new
      simulations had to be done. \\      
    \end{abstract}
  \end{@twocolumnfalse}
  ]

@@

\section{INTRODUCTION}
On March 7-10 of 2017, a 5 $\mu$A electron beam was delivered to a BeO and Carbon targets for the first time to experimental
Hall C since the 12 GeV upgrade. The beam was delivered as part of the Key Performance Parameters (KPP) required by the
Department of Energy (DOE) to demonstrate the operability of the High Momentum Spectrometer (HMS) and Super HMS (SHMS). 
Hall C was able to demonstrate KPP in four days of beam time before an important component of the accelerator was damaged which
caused to accelerator to shut down for repair. The accelerator is expected to be operational starting December 4, 2017. As a
result of this delay, the commissioning experiments that were scheduled to run on Fall 2017 have now shifted to Spring 2018.
This time window has allowed the Hall C collaboration to work extensively in preparation for the commissioning of the
spectrometers on December. \\
\indent One of the projects I have been involved in is the ongoing work on testing and commissioning the 12 GeV HMS Drift
Chambers. The chambers were constructed at Hampton University by Dr. Liguang Tang and his graduate students in 2016.
They were made the same design as the SHMS chambers, but slightly different size. The chambers were transported to Jefferson Lab on November 2016,
where they underwent extensive tests as part of conditioning the chambers to sustain High Voltages using a gas mixture\footnote{This gas mixture is non-flammable and
at lower cost compared to the gas mixture that the chambers run on during an experiment, which is why it is preferred during the testing phase of the detector.}
of 75:25 Argon/CO$_{2}$ by volume. The chambers were found to be operational below 1850 V which is below the expected value\footnote{The HMS chambers
are the same design as the SHMS chambers, which operate at $\sim$ 1940 V.} of 1940 V. At high voltages above 1850 V, the chambers drew a significant amount of
dark current which made the signals from the chamber noisy.
It was determined that the most likely cause of the large currents drawn was the gas mixture[Ar/CO$_{2}$] being used, so the one of the chambers (HMS DC II) was transported to the
experimental Hall C where a gas mixture of 50:50 Argon/Ethane by volume was used. A test stand for the chamber was set up in the HMS hut, where it has been tested and
verified to be operational with the new gas mixture. The other chamber (HMS DC I) exhibited similar symptoms as the first with the addition that it had a few missing channels
due to a bad connection between the sense wires and the discriminator cards. The second chamber is now ready to be transported to Hall C for further tests with the
Argon/Ethane gas mixture before it can be put in the detector stack.\\
\indent A second project I am currently involved in is the determination of electronic and computer dead times. In nuclear/particle physics experiments,
the number of physics events (triggers) are counted via nuclear electronic modules. These triggers ultimately get processed by the DAQ before being
written to tape. The deadtime refers to the time window in which the modules are unable to process triggers and physics events are lost.
The electronic deadtime contribution comes from the electronic modules having a maximum rate capability. Typically,
the electronic modules in Hall C can achieve rates from a few MHz to few hundred MHz. Once the modules reach their limit, a pileup of signals can occur which
contributes to the total dead time. The computer deadtime contribution comes from finite processing time of the DAQ. These rates are typically on the order of
a few kHz, therefore, the dominant contribution of total dead time comes from the DAQ, since it takes a few kHz of data before physics events
are lost. These measurements are important for the determination of high precision cross section measurements in Hall C, since knowing how many events
are actually lost can make a significant difference in the uncertainty of the cross section. \\
\indent My thesis experiment, the \textit{Electro-Disintegration of Deuteron at Very High Missing Momenta} is projected to run towards the end of
February 2018, and will receive a total of six days of beam time. The experiment will be done at four different spectrometer configurations. The kinematic
setting have changed slightly from the original proposal, and will be briefly discussed in this paper.

\section{12 GeV HMS Drift Chambers}

\subsection{Design and Operation of the Chambers}
The new HMS Drift Chambers were designed to be geometrically the same as the SHMS Drift Chambers. Each chamber consists of 6 wire planes and each wire
plane is located between two cathode planes. The wire planes consist of alternating field and sense wires. The U, U', V and V' planes consist of
96 sense wires each and are oriented 60$^{o}$ relative to the +x-coordinate. The X and X' planes consist of 102 sense wires and are oriented perpendicular to the x-axis (See
Figures \ref{fig:dc1_planes} and \ref{fig:dc_wires}).
\begin{figure}[h!]
  \centering
  \includegraphics[width=3.0in, height=2.5in]{dc2_tests/HMS_DC1_Planes.pdf}
  \caption{HMS Drift Chamber 1 wire planes.}
  \label{fig:dc1_planes}
\end{figure}
\begin{figure}[h!]
  \centering
  \includegraphics[width=3.0in, height=1.5in]{dc2_tests/HMS_DC_Wires.pdf}
  \caption{HMS Drift Chamber wire orientation for planes a) U, U', b) V, V' and c) X, X' where
  the direction of the beam (+z) is into the plane of the paper.}
  \label{fig:dc_wires}
\end{figure}
The cathode planes and field wires are held at a negative high voltage and the sense wires are grounded which establihses a potential gradient causing an electric field
between high voltage and grounded wires. Each chamber is filled with a gas mixture\footnote{Argon is mixed with CO$_{2}$ or Ethane, where Argon is the ionizing agent the
particles interact with, and Ethane/CO$_{2}$ are the quenching elements to control avalanche produced by secondary ionization} of Argon with either CO$_{2}$ or Ethane.
As the particles traverse the chamber, the free electrons from ionized Argon drift towards the sense wires producing a detectable signal which is read out via discriminator
cards and into electronic modules.
